LCA in R using tidySEM

### 2022-11-22 example_analysis/LCA_Tutorial.Rmd
Update November 23
- updated the literature list
- added use case examples
- corrected spelling/grammar mistakes
- adjusted spacing, also in the tutorial

*** things to resolve:
1) Caspar should add Well-being compass study to Zotero and update the in-text citation for that use case
2) the plot_density function doesn't work, also the example in the documentation;
I would like to show how to use it in class enumeration section of the tutorial
3) final proof-reading, anything to add?

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent class
  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

```{r load_packages, include = FALSE}
library("papaja")
#bibliography      : ["r-references.bib"]
library(worcs)
run_everything = FALSE
```

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. @vermunt_latent_2004 defined LCA more generally as
virtually any statistical model where "some of the parameters [...]
differ across unobserved subgroups".

Despite its popularity, there is a lack of standards for estimating and
reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
reporting guidelines for a specific type of LCA known as latent growth
models, general reporting guidelines for all LCA are still lacking. This
complicates manuscript review and assessment of the quality of published
studies, and introduces a risk of misapplications of the technique. The
present paper seeks to address this gap in the literature by suggesting
updated guidelines for estimation and reporting of LCA, based on current
best practices. Importantly, in order to lower the barrier of entry and
ensure reproducibility of all examples, this paper exclusively relies on
free, open source software in R. Our goal is to make best-practices in
LCA widely accessible.

## Defining Latent Class Analysis

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership. The concept of LCA can be understood in different ways. A
mixture model assumes that the study population is composed of $K$
subpopulations or classes. It further assumes that the observed data are
a mixture of data generated by class-specific models. The simplest
univariate model is a normal distribution, which can be described with
two parameters: the mean and the variance. Commonly, the same model is
estimated across all classes, but with different parameters for each
class (i.e., class-specific means and variances). Mixture modeling then
estimates both the parameters for each class, and the probability that
an individual belongs to each class. This way, describing the whole
sample with a single mean and variance is a special case where we assume
the sample is homogeneous in function of these parameters.

As an illustrative example, imagine that a detective wants to know if it
would be possible to use mixture modeling to identify the sex of a
suspect, based on footprints found at the crime scene. To test the
feasibility of this approach, the detective records the shoe sizes and
sex of 100 volunteers. The resulting observed data look like this:

```{r echo = FALSE}
library(tidySEM)
library(ggplot2)
set.seed(1)
n = 100
C <- sample(c("Man", "Woman"), n, replace = TRUE)
means <- c(Man = 9, Woman = 7)
X <- rnorm(n, mean = means[C], sd = 1)
X <- round(X*2)/2
obsparam <- tapply(X, C, mean)
shoesize_shapiro <- shapiro.test(X) # Shapiro-Wilk test of normality
```

```{r echo = FALSE, eval = run_everything}
est <- mx_profiles(data.frame(X), 2)
estparam <- table_results(est, c("label", "est", "matrix"))
estparam <- as.numeric(estparam$est[estparam$matrix == "M"])
props <- class_prob(est, "individual")$individual[ , "predicted"]
dput(props, "props.txt")
dput(estparam, "estparam.txt")

p <- ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
ggsave("shoedens.png", p, device = "png")
```

```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
props <- eval(parse("props.txt"))
estparam <- eval(parse("estparam.txt"))
knitr::include_graphics("shoedens.png")
```

The distribution is evidently bimodal, which bodes well for the intended
mixture model. In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that
the model estimates the mean shoe size of the two groups are equal to
`r report(estparam[1], equals = FALSE)` and
`r report(estparam[2], equals = FALSE)`, which is close to the true
means of the two groups, namely `r report(obsparam[1], equals = FALSE)`
and `r report(obsparam[2], equals = FALSE)`. When tabulating estimated
group membership against observed (known) group membership, it can be
seen that women are classified with a high degree of accuracy, but men
are not:

```{r tabshoe, results="asis"}
library(kableExtra)
tab <- table(C, props)
tab <- data.frame(Observed = rownames(tab), as.data.frame.array(tab))
names(tab)[2:3] <- c("Class 1", "Class 2")
rownames(tab) <- NULL
kbl(tab, caption = "Observed group membership by estimated class membership.")
```

LCA is like confirmatory factor analysis, except that the continuous
latent variable is substituted with a categorical latent variable. Both
assume that the indicators are related, and these relations can be
represented by the underlying latent variable. One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes.While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but is
instead estimated -- with measurement error -- based on the data.
Whereas in a multi-group model, the data are split by group and treated
as independent samples, in LCA, all cases contribute to the estimation
of all parameters in all groups. The relative contribution of each case
to the parameters of each group is determined by that case's posterior
probability of belonging to that group.

In factor analysis, the latent variable is continuous, so the relation
between indicators is defined by the communalities between them, and the
latent variable represents the error corrected approximation of the
sample. While in LCA, the relation between indicators is define by how
well they separate the classes, as a multinomial regression between
indicators and latent variable [@masyn_latent_2013]

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm. In line with this
perspective, LCA is sometimes described as "model-based clustering"
[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model. Specifically, the
posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data
under given the class-specific model. For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for. This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error. 0

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable --
true class membership -- is not known, and the term *classification*
refers to the fact that the algorithm is predicting a categorical
outcome -- class membership.

## A Taxonomy of Latent Class Analysis Methods

In this paper, we use the term latent class analysis to refer to a
family of techniques that estimate latent class membership based on a
parametric model of observed indicators. From a historical perspective,
the term was initially conceived to refer to analyses with categorical,
usually binary indicators [@vermunt_latent_2004]. Nowadays, there are
a number of related techniques, known by distinct names, that serve a
similar purpose. The term "latent class analysis" seems most appropriate
as an umbrella term for this broader class of models, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators. Given the
abundance of terms in use for closely related classes of models, we will
provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes [@collins_latent_2009].

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009]. As an example, a LPA of ocean
plastic particles found two classes of particles based on length and
width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent
indicators. This means that, within each class, one or more continuous
latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these
continuous latent variables. A common application of this approach is in
longitudinal research, where the indicators reflect one construct
assessed at different time points. Examples of this approach include
*growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe grouped
trajectories over time. The growth mixture model is a latent class model
where the parameters that indicate class membership are the intercepts
and variances, and typically covariances of the latent growth variables,
e.g., a latent intercept and slope. This technique assumes that
individuals within a class can have heterogenous trajectories. If the
variance of the growth parameters is fixed to zero, it is known as a
latent class growth analysis. This latter approach assumes that all
individuals within a class share the same identical trajectory, and that
any variance in the indicators not explained by the class-specific
latent trajectories is due to residual error variance.
<!-- MGV: some reference for this description -->

The term latent class analysis originally referred to cases where the
observed indicators were categorical [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*. Latent class models
with ordinal indicators are parameterized differently from mixture
models. One common parameterization assumes that each categorical
variable reflects an underlying standard normal distribution. The
parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.

This paper will primarily focus on mixture models and special cases
thereof, although most of the suggested guidelines are applicable to all
LCA methods.

## Use Cases for Latent Class Analysis

<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->

There are several use cases for which LCA methods are suitable. One
example is to test a theory that postulates the existence of a
categorical latent variable. For example, *identity status theory*
posits that, at any given point in time, adolescents reside in one of
four identity statuses. Latent class analysis can be used to identify
these four statuses based on observed indicators (e.g., self-reported
identity exploration and commitment). If results indicate that the data
are better described by a different number of classes, or that the
four-class solution does not correspond to the predicted pattern of
responses on the indicators, then the theory may be called into
question.

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be used as a diagnostic
aid when several clinical indicators can be used to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental
[@wu_abuse_2011] health problems, or health outcomes after recovering from cancer [@hardy_exploring_2018]. The example of shoe size is a
rudimentary illustration of this type of application.

LCA can be used as a descriptive analysis where a researcher wishes to
describe their sample and identify a few prototypes based on many
variables. As an example, if a survey among all Dutch academics was
carried out with the goal of bringing about a funding reform, LCA could
be used to discover the types of publications that get funding.

With LCA, our goal could be to inductively identify the number of
classes. For instance, if we believe that a variable represents a group,
but don't know how many groups there are, LCA may be an appropriate
technique to answer this question. For example, @hopfer_social_2014
studied substance use, sexual behavior, and mental health status of
urban population in Winnipeg, Canada. The underlying assumption was that
there were different risk profiles, but their number was not known. From
a collection of indicators, LCA provided evidence that there are four
distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals. In a peer
harassment study, @giang_using_2008 used LCA to classify over 2,000
sixth grade students into aggressor and victim latent classes. The
five-class solution comprised of classes of victims, aggressors, and
socially adjusted students. For instance, it revealed that there were
two types of victims: highly-victimized aggressive-victims and
highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture
classes well. High quality indicators will be strongly related to the
latent variable and will lead to good class separation. This
relationship of high quality indicators to the latent variable is
reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on
LCA, see @geiser_is_2014. Therefore, one could use conditional response
probabilities for each item to assess its quality with regards to how
well it helps separate the latent classes. From this, a theory about the
selection of indicators could be informed.

An extension of LCA is that containing covariates which can be used to
predict class membership. In this approach, we not only model the latent
class variable based on indicators, but we also relate the class
membership to other explanatory variables (@vermunt_latent_2017). An
example of using covariates comes from @nozadi_moderating_2016 who
applied LCA to identify the probability of children's membership to an
anxiety class. The authors tested several covariates including
children's age, sex, and accuracy scores. Age and sex were not found to
be related to the children's latent class membership, hence these
covariates were excluded from the analysis. Accuracy scores, however,
were related to probabilities of being in anxiety and attention-anxiety
classes. Therefore this covariate was kept as a valuable predictor.

When our interest is the prediction of one or more outcomes, LCA can be
used to construct latent classes as categorical predictors.
@lanza_latent_2013 demonstrated how LCA can be used to classify
adolescents into depression classes, and subsequently these classes can
be used to predict smoking, grades, and delinquency. The study showed
that the outcomes predicted by class membership can be binary (regular
smoking), continuous (grades) or count (delinquency).

These mixture models can also be useful to evaluate quality of measurement across instrument. For example, employment status was collected across register and survey data, and there is disagreement between them. You could think that register data is without error, but it has been shown that is victim of measurement error as well. In this case mixture models was used to evaluate the underlying factor of "error corrected employment status", by using two indicators of the same question and identify random and systamtic measurement error in either indicator [@pavlopoulos_measuring_2015].

In addition to these applications, LCA can be used for dimensionality
reduction as the resulting groups summarize response patterns on a large
number of indicators. For example, @macgregor_symptom_2021 investigated
symptom profiles among injured U.S. military personnel. They used
fifteen dichotomous items from the Post-Deployment Health Assessment
survey as LCA indicators. Combinatorics informs us that fifteen
dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a
method of dimensionality reduction. A five class solution was found to
have the best fit according to both statistical criteria and clinical
interpretability.

<!-- CJ: (on the next paragraph)  -->

<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->

<!-- That is not what this paragraph is about.  -->

<!-- This paragraph is about motivations for using LCA. -->

<!-- One motivation is if your data violates multivariate normality -->

<!-- because violating multivariate normality means that you have multiple groups -->

<!-- and you can use LCA to extract those. That is the only assumption  -->

<!-- we have to discuss here, we just need to explain it. -->

Finally, LCA can be used to deal with data which violate certain
assumptions. As discussed in the shoe size example, LCA can deal with
violations of normality. In fact, LCA assumes the population
distribution is a non-normal mixture of $K$ normal distributions, and it
can discover the value of $K$, i.e. generate a $K$-class solution.

LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question.

# Best Practices

## Best Practices in Estimation

The best practices in estimation, as outlined in Table
\@ref(tab:checkest), are rooted in existing recommendations for best
practices for estimating specific subtypes of LCA, including latent
class growth analysis [@van_de_schoot_grolts-checklist_2017] and latent
class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
These were generalized to be relevant to all types of LCA, and updated
to current best practices, as explained below.

```{r checkest, include=TRUE}
tab_check <- data.frame(Item = c(
  "Examining Observed Data",
  "Handling Missing Data",
  "Alternative Model Specifications",
  "Software",
  "Algorithm",
  "Class Enumeration",
  "Model Fit Indices",
  "Classification Diagnostics",
  "Interpreting Class Solutions",
  "Label switching"

))
tab_check$`#` = paste0(1:length(tab_check$Item), ".")
kbl(tab_check[, c("#", "Item")])
```

### Examining Observed Data

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection. Special attention should be paid to level of
measurement of the indicators. Finite Gaussian mixture models (including
LPA) are only suitable for continuous variables. Indicators with an
ordinal level of measurement are likely to violate the assumption of
within-class normal distributions of mixture models [see
@vermunt_k-means_2011]. Personal experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
indicators are treated as continuous [e.g., @norman_likert_2010] or 
with 5+ categories [@rhemtulla_when_2012], this
certainly does not imply that all methods are. It is certainly unlikely
that such ordinal variables can be treated as a *mixture* of multiple
normal distributions. The problem becomes egregious when the number of
classes estimated equals or exceeds the number of categories; in this
case, each class-specific mean could describe a single response
category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture
models is that observed indicators are normally distributed
[@spurk_latent_2020]. This is incorrect. When the number of classes is
greater than one, Gaussian mixture models assume that the observed indicators are
a mixture of multiple (multivariate) normal distributions. In our shoe
size example, it can be seen that the population distribution is
comprised of two normal distributions. When examined visually, the
population distribution is evidently bimodal. The Shapiro-Wilk normality
test ($W = 0.971$, $p < 0.05$ ) rejects the null hypothesis that the
sample comes from a normally distributed population. Yet, this is a
prototypical example of a mixed population distribution where LCA can
discover latent groups. If the population distribution were instead
normal, there would be no classes to extract as the whole population
would belong to a single (homogeneous) class.

Extensive descriptive statistics (including the number of unique values,
variance of categorical variables, and missingness; see next subsection)
can be obtained using the function `tidySEM::descriptives(data)`. Note,
however, that sample-level descriptive statistics are of limited value
when the goal of a study is to identify subsamples using latent class
analysis. Plots (density plots for continuous variables, and bar charts
for categorical ones) may be more diagnostic. Note that density plots
can also aid in the choice of the number of classes, as further
explained in the section on visualization. Descriptive statistics and
plots can be relegated to online supplements, provided that these are
readily accessible [consider using a GitHub repository as a
comprehensive public research archive, as explained in
@van_lissa_worcs_2021].

### Handling Missing Data

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; @enders_applied_2022]: Missing
completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
missing at random (MAR), which means that missingness is contingent on
the *observed* data (and can thus be accounted for); and finally missing
not at random (MNAR), which means that missingness is related to
unobserved factors. It is possible to conduct a so-called "MCAR" test,
for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
note that the name "MCAR test" is somewhat misleading, as the
null-hypothesis of this test is that the data are not MAR, and a
significant test statistic indicates that missingness is related to the
observed data (MAR). A non-significant test statistic does not
distinguish between MCAR or MNAR. As Little's classic MCAR test relies
on the comparison of variances across groups with different patterns of
missing data, it assumes normality [@little_test_1988]. This assumption
is tenuous in the context of LCA. A non-parametric MCAR test, as
provided by Jamshidian and Jalal, may be more suitable
[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
central R-repository CRAN due to lack of maintenance. For this tutorial,
I have re-implemented it in the `mice` package as `mice::mcar()`, with a
fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due diligence, it is
important to emphasize that missingness is adequately handled by default
in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` which is
the backend of `tidySEM`. These packages use Full Information Maximum
Likelihood (FIML) estimation, which makes use of all available
information without imputing missing values. FIML is a best-practice
solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
[@lee_comparison_2021]. FIML estimation assumes that missingness is
either MCAR or MAR. Thus, one would typically proceed with FIML,
regardless of the outcome of an MCAR test. Although FIML does not, by
default, handle missingness in exogenous variables -- all indicator
variables in LCA are endogenous, so this is not a concern.
<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->

Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally expensive, and conducting
them on multiple imputed datasets may be unfeasible. Second, because
there is no straightforward way to integrate LCA results across multiple
datasets. To conclude; our recommendation is to inspect missingness
(e.g., using `mice::MCAR()`) and report the proportion of missingness
per variable (e.g., using `tidySEM::descriptives()`), before proceeding
with FIML. One minor concern is that the K-means algorithm, which
`tidySEM` uses for determining starting values, is *not* robust to
missing values. When it fails, `tidySEM` automatically switches to
hierarchical clustering, unless the user specifies a different
clustering algorithm or uses manual starting values.

### Alternative Model Specifications

In order to aid researchers working with latent trajectory models,
@van_de_schoot_grolts-checklist_2017 developed a protocol called
Guidelines for Reporting on Latent Trajectory Studies (GRoLTS). Two of
the GRoLTS guidelines (namely, 6a and 6b) refer to considering
alternative model specifications. Both are discussing specific cases in
latent trajectory model specification. The first is about whether the
variance of the growth parameter is estimated freely or fixed, and the
second is about whether conditional independence is assumed, as the
researcher might also want to free-up the variance-covariance structure.
In LCA, there are also many different ways to specify the model. Means,
variances, and covariances between the indicators can either be
constrained across classes, or estimated freely. Researchers using LCA
should transparently report their chosen parametrization as well as
discuss different parametrizations that were tested in the process.

Different types of latent class models have different parameters. For
example, mixture models and latent profile analyses typically have
class-specific means, variances, and covariances. Latent growth analyses
have the same parameters, but with respect to the latent growth
variables. Latent class analyses with ordinal indicators have
thresholds. All of these parameters can be freely estimated, constrained
to be equal across classes, or fixed to a certain value (e.g., to zero).
The total number of parameters thus scales with the number of estimated
classes. Consequently, LCA methods have a potentially very high number
of parameters. As any of these parameters could be misspecified, it is
important to consider alternative model specifications. However,
alternative model specifications may be approached differently depending
on whether an analysis is data driven (exploratory), or theory driven
(confirmatory). This distinction has remained underemphasized in prior
writing.

Prior literature on LCA has emphasized its exploratory applications
[@nylund_deciding_2007]. In exploratory LCA, a large number of models
are typically estimated in batch, with varying numbers of classes and
model specifications. The "correct" model specification is then
determined based on a combination of fit indices, significance tests,
and interpretability. For latent profile analysis, the function
`tidySEM::mx_profiles(classes, variances, covariances)` largely
automates this process. The argument `classes` indicates which class
solutions should be estimated (e.g., 1 through 6). The argument
`variances` specifies whether variances should be `"equal"` or
`"varying"` across classes. The argument `covariances` specifies whether
covariances should be constrained to `"zero"`, `"equal"` or `"varying"`
across classes. The means are free to vary across classes by default,
although the more general function `tidySEM::mx_mixture()` could be used
to circumvent this. After all models have been estimated, the function
`tidySEM::table_fit()` can be used to obtain a model fit table suitable
for determining the optimal model according to best practices. Note,
however, that this table does not include the bootstrapped likelihood
ratio test (BLRT) by default, because this test is very computationally
expensive. It is recommended to use the function `tidySEM::BLRT()` to
compare a shortlist of likely candidate models based on other fit
indices. More on fit indices can be found in the Model Fit Indices
subsection of this paper.

Confirmatory LCA typically requires less comprehensive alternative model
specifications. For example, in the context of a preregistered analysis,
the main models of interest would have been specified a priori. Even in
this case, the theoretical model could be compared to a few others to
contextualize it.

```{r, include = FALSE}
grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
"Is information presented about the mean and variance of time within a wave?", 
"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
"Is a description provided of how missing data in the analyses were dealt with?", 
"Is information about the distribution of the observed variables included?", 
"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs. LGMM) and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative shape/functional forms of the trajectories described?", 
"If covariates have been used, can analyses still be replicated?", 
"Is information reported about the number of random start values and final iterations included?", 
"Are the model comparison (and selection) tools described from a statistical perspective?", 
"Are the total number of fitted models reported, including a one-class solution?", 
"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
"If classification of cases in a trajectory is the goal, is entropy reported?", 
"Is a plot included with the estimated mean trajectories of the final solution?", 
"Are plots included with the estimated mean trajectories for each model?", 
"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
```

### Software

Many software packages are available for the estimation of LCA models.
Some of these packages have limited functionality, or implement specific
innovations. Other packages implement LCA in the context of a more
flexible structural equation modeling framework. The most notable
examples of the latter are the commercial programs Mplus and Latent
GOLD, and the free open source R-package OpenMx. The commercial packages
stand out because they offer relatively user-friendly interfaces and
implement sensible defaults for complex analyses, including LCA methods.
This lowers the threshold for applied researchers to adopt such methods.
Commercial software also has several downsides, however. One such
downside is that use of the software is restricted to those individuals
and institutions who can afford a license. Another downside is that the
source code, being proprietary, cannot be audited, debugged, or enhanced
by third parties. This incurs the risk that mistakes in the source code
may go unnoticed, and curbs progress as software developers cannot add
new functionality.

Conversely, the free open source program OpenMx [@neale_openmx_2016] is very flexible, but
not very user-friendly. We directly address this limitation using the
`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
threshold for latent class analysis using `OpenMx`. It adheres to best
practices in estimation and reporting, as described in this paper. The
user interface is simple, making use of the model syntax of the widely
used `lavaan` R-package [@rosseel_lavaan_2012]. This syntax offers a human-readable way to
specify latent variable models. Minor enhancements are made to simplify
the specification of LCA.

Because of the limitations of the aforementioned tools, we set out to
develop a free tool that provides sensible defaults and is easy to use,
but provides the option to access and modify all of the model inputs
(i.e., low barrier, high ceiling). `tidySEM` interfaces with existing
tools, and is able to translate between what existing tools are capable
of and what researchers and analysts carrying-out person-oriented
analyses would like to specify. Furthermore,`tidySEM` facilitates
fully-reproducible analyses and contributes to open science.

### Algorithm

LCA parameters and model fit statistics can be estimated in a variety of
ways. The choice of the estimator depends on the presence of missing
values, sample size, number of indicators, and available computational
resources [@weller_latent_2020]. A commonly used technique is maximum
likelihood (ML) estimation with the expectation-maximization (EM)
algorithm as a local optimizer. Imagine we are estimating two
parameters, e.g. the class-specific means $\mu_c$ on a continuous
indicator (ignoring the variance for now). The EM algorithm will attempt
to find a combination of values for these two parameters that maximizes
the likelihood ($LL$) of all observed data. In practice, instead of
maximizing $LL$, often $-2*LL$ is minimized, as this offers
computational advantages. We can think of this optimization problem as a
three-dimensional landscape: The X and Y dimensions are determined by
the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the
Z-dimension is determined by $Z = -2*LL$. The optimizer must find the
deepest "valley" in this landscape, which reflects the combination of
$\mu_1$ and $\mu_2$ that maximizes the likelihood of the data. The EM
optimizer behaves somewhat like a marble dropped in this landscape. It
is dropped at some random point in space, and will roll into the nearest
valley. The problem is that, once EM rolls into a valley, it will settle
on the bottom of that valley (this is known as "convergence"). It cannot
climb out again. Thus, if their are multiple valleys, the risk is that
the optimizer gets stuck in a shallower valley (a "local optimum"), and
never discovers the deepest valley (the "global optimum", or best
solution). One solution to this problem is to drop many marbles at
random places, compare their final $-2*LL$ values, choose the solution
with the lowest $-2*LL$, and make sure that several marbles replicated
this solution. This is the "random starts" approach.

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values, many of the
marbles are likely to be very far away from a "good enough" solution.
Two innovations may improve the estimation procedure. The first is that,
instead of picking random starting values, a "reasonable solution" may
be used for the starting values. For example, if we assume that the
different classes are likely to have different mean values on the
indicators, then the K-means clustering algorithm can be used to
determine these cluster centroids. We can compute the expected values of
all model parameters by treating the K-means solution as a known class
solution, and use these as starting values for a mixture model. One
remaining concern is that this approach may result in starting values
close to a local optimum, and that the EM algorithm will thus never find
the global optimum. A second innovation addresses this concern. Instead
of using EM, it is possible to use an optimizer that can climb out of a
valley. Simulated annealing iteratively considers some "destination" in
the landscape, and compares its likelihood to the current one. If the
destination likelihood is higher, the estimator moves there. If the
destination likelihood is *lower*, the estimator still moves there
occasionally, based on probability. This latter property allows it to
escape local optima, and find the global optimum.

By default, `tidySEM` derives starting values using K-means clustering,
and identifies the global optimum solution using simulated annealing.
Once a solution has been found, simulated annealing is followed up with
a short run of the EM algorithm, as EM inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors. Note that these defaults can be manually overridden.

A recent paper suggested maximum likelihood with robust standard errors
should be used when the observed indicators are not normally distributed
[@spurk_latent_2020]. This statement is incorrect, and may lead readers
to believe that they must use commercial software, as robust maximum
likelihood is currently only implemented in Mplus and LatentGOLD. As
explained before, mixture modeling assumes that observed data are a
mixture of (multivariate) normal distributions; thus, the observed
indicators will likely not be normally distributed.
<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->

### Class Enumeration

As explained, LCA can be done in an exploratory or in a confirmatory
fashion. In exploratory LCA, a sequence of models is fitted to the data
with each additional model estimating one more class than the previous
model. These models are then compared and the best solution is selected
as the final class solution. In some cases, prior theory can inform the
researcher about the number of classes to expect. Even in such
confirmatory LCA cases, it is nonetheless useful to know if the
theoretical model is markedly better than those with differing numbers
of classes. Therefore, it may always be useful to compare different
class solutions.

>>>>

---
title: "LCA in R using tidySEM"
output: html_notebook
---
# Tutorial
